---
title: 'Homework 4: Tidy data and joins'
author: "Santiago David"
date: '2017-10-06'
output: github_document
---

### Load data and packages
```{r}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(gapminder))
```

## Data reshaping

### Activity #2

**Objective**: Make a tibble with one row per year and columns for life expectancy for two or more countries, and use that to make a plot.

**Process**: Colombia and Venezuela were the same country many years ago, I'm curious how life expectancy have changed in their recent past (as separate countries), so I will analyze them.
First, I filter gapminder by those two countries and all years. Then, I select the variables `country`,`year`, and `lifeExp`, and use `spread` to convert that table into its wide format, and save as a new object.

```{r}
colven <- gapminder %>% 
  filter(country %in% c("Colombia", "Venezuela")) %>% 
  select(country, year, lifeExp) %>% 
  spread(country, lifeExp)
knitr::kable(colven, col.names = c("Year", "Life expectancy in Colombia", "Life expectancy in Venezuela"))
```

**Figure**: We can use this new data shape to look at the comparison of life expectancy of Colombia vs Venezuela. I will include a `geom_abline` with slope=1 and intercept=0, so we can see how different is this comparison to a perfect 1 to 1 relationship in life expectancy for both countries over the years.

```{r}
colven %>% 
  ggplot(aes(x = Colombia, y = Venezuela)) +
  geom_point() +
  geom_label(aes(x = Colombia, y = Venezuela, label = year, vjust = -0.5)) + #add years to points
  ylim(50, 75) + xlim(50, 75) +
  geom_abline(slope = 1, intercept = 0, colour = "4", linetype="longdash") + #add line
  ggtitle("Comparison of Life Expectancy between Venezuela and Colombia") +
  theme_bw()
```

**Observations**: From this figure and table we can see that both countries have increased their life expectancy over the years, but Venezuela has always had a life expectancy higher than Colombia (above blue line), especially in the 50's-60's. Both countries have a more similar life expectancy in recent years (closer to blue line).

**Additional resources**: I got help from the ggplot2 reference webpage for adding [text and labels to a plot](http://ggplot2.tidyverse.org/reference/geom_text.html).


## Join, merge, look up

### Activiy #1

**Objective**: Create a second data frame, complementary to Gapminder. Join this with (part of) Gapminder using a dplyr join function and make some observations about the process and result. Explore the different types of joins.

**Process**: Rather than creating a second data frame, I decided to get some open access data from the web to practice `joins` for this activity. I'm interested in global land use change, so I googled and found open access data for Agricultural land (sq. km) for all countries in the world from [The World Bank](http://www.worldbank.org) official webpage. The data can be downloaded as .csv file from [this site](https://data.worldbank.org/indicator/AG.LND.AGRI.K2?end=2016&start=1961&view=chart), and because the license says "open", I'm assuming it is okay to use it here as an example (not 100% sure). 

Loading the data
```{r}
agrland <- read.csv(file.choose("API_AG.LND.AGRI.K2_DS2_en_csv_v2.csv"))
str(agrland)
as.tibble(agrland)
View(agrland)
```


```{r}
View(gapminder)
```

Temporal
In case I decided to use it, this is the webpage for land use and agricultural land for the world countries (https://data.worldbank.org/indicator/AG.LND.AGRI.K2?end=2016&start=1961&view=chart)


## But I want to do more!

I also tried the activity #3 but after thinking for a while, I dont see how this new shape of the data will help us to produce a better visualization. As we know, ggplot2 prefer long format, so other than the sort of plot produced in the first part of this homework to compare two countries, I'm not sure this table will help. I would still prefer a "wide" table vs a "long" one for visual comparison of data.

```{r}
life_exp_mean <- 
  gapminder %>% 
  group_by(continent, year) %>% 
  summarise(life_mean = mean(lifeExp)) %>% 
  spread(continent, life_mean)
knitr::kable(life_exp_mean)
```