---
title: "Gapminder exploration with dplyr and ggplot2 hw02"
output: github_document
---

## Bring data in

We already installed the packages, so I'll just load them
```{r}
library("gapminder")
library("tidyverse")
```

## Smell test the data

Is it a data.frame, a matrix, a vector, a list?
```{r}
typeof(gapminder)
```
**Answer**: Gapminder is an object of the type *list*. I used the function `typeof` to ask for its type

What’s its class?
```{r}
class(gapminder)
```
**Answer**: If I understand *data structures*, gapminder is both a tibble and a dataframe as we can see using `class` for the different elements in the list.

How many variables/columns?
```{r}
ncol(gapminder)
```
**Answer**: There are 6 columns in gapminder. The easiest way to see this is using `ncol` as well as for the number of rows using `nrow`

How many rows/observations?
```{r}
nrow(gapminder)
```
**Answer**: There are 1704 rows

Can you get these facts about “extent” or “size” in more than one way? Can you imagine different functions being useful in different contexts?

**Answer**: Yes! I'm sure there are tons of ways to do that in r. Two ways I can think of are using `dim` to see the dimensions of the dataframe, and `str` to see the whole structure.
Personally, I always use `str` to inspect not only the dimensions of the dataframe but also the type of all variables. But in a different context if someone is only interested in the size of the dataframe, `dim` would make the trick.
```{r}
dim(gapminder)
str(gapminder)
```

What data type is each variable?

**Answer**: We can easily check this using `str` as seen above. *country* and *continent* are objects of the type `Factor`, *year* and *pop* are `integer`, and *lifeExp* and *gdpPercap* are `numeric`.


## Explore individual variables
I will explore the categorical variable `country` and two quantitative variables population `pop` and year `year`

What are possible values (or range, whichever is appropriate) of each variable?

**Answer1**: For the categorical variable, we can see all possible values or *non-repeated* entries using the function `unique`

```{r}
unique(gapminder$country)
```

We can also see from this list of names that there are 142 countries. Or we can directly ask that using `n_distinct` or `length`
```{r}
length(unique(gapminder$country))
```

For the **quantitative variables**, I am assuming there wont be data for many years so we can also use `unique` to see the specific years and `range` to see the oldest and most recent. But for population, we can expect that all countries will have different pop sizes every year, so that would be a very long list, in this case it is better to explore the range.

For `year`:
```{r}
unique(gapminder$year)
range(gapminder$year)
n_distinct(gapminder$year)
```

For `pop`
```{r}
range(gapminder$pop)
```

**Answer2**: There are data from 12 years starting in 1952 with the most recent data from 2007.
There are population sizes starting from 60011 to 1318683096 (individuals!, I guess...)


What values are typical? What’s the spread? What’s the distribution? Etc., tailored to the variable at hand
