---
title: "Gapminder exploration with dplyr and ggplot2 hw02"
output: github_document
---

## Bring data in

We already installed the packages, so I'll just load them
```{r}
library("gapminder")
library("tidyverse")
```

## Smell test the data

Is it a data.frame, a matrix, a vector, a list?
```{r}
typeof(gapminder)
```
**Answer**: Gapminder is an object of the type *list*. I used the function `typeof` to ask for its type

What’s its class?
```{r}
class(gapminder)
```
**Answer**: If I understand *data structures*, gapminder is both a tibble and a dataframe as we can see using `class` for the different elements in the list.

How many variables/columns?
```{r}
ncol(gapminder)
```
**Answer**: There are 6 columns in gapminder. The easiest way to see this is using `ncol` as well as for the number of rows using `nrow`

How many rows/observations?
```{r}
nrow(gapminder)
```
**Answer**: There are 1704 rows

Can you get these facts about “extent” or “size” in more than one way? Can you imagine different functions being useful in different contexts?

**Answer**: Yes! I'm sure there are tons of ways to do that in r. Two ways I can think of are using `dim` to see the dimensions of the dataframe, and `str` to see the whole structure.
Personally, I always use `str` to inspect not only the dimensions of the dataframe but also the type of all variables. But in a different context if someone is only interested in the size of the dataframe, `dim` would make the trick.
```{r}
dim(gapminder)
str(gapminder)
```

What data type is each variable?

**Answer**: We can easily check this using `str` as seen above. *country* and *continent* are objects of the type `Factor`, *year* and *pop* are `integer`, and *lifeExp* and *gdpPercap* are `numeric`.


## Explore individual variables
I will explore the categorical variable `country` and two quantitative variables population `pop` and year `year`

What are possible values (or range, whichever is appropriate) of each variable?

**Answer1**: For the categorical variable, we can see all possible values or *non-repeated* entries using the function `unique`

```{r}
unique(gapminder$country)
```

We can also see from this list of names that there are 142 countries. Or we can directly ask that using `n_distinct` or `length`
```{r}
length(unique(gapminder$country))
```

For the **quantitative variables**, I am assuming there wont be data for many years so we can also use `unique` to see the specific years and `range` to see the oldest and most recent. But for population, we can expect that all countries will have different pop sizes every year, so that would be a very long list, in this case it is better to explore the range.

For `year`:
```{r}
unique(gapminder$year) # only unique entries
range(gapminder$year) # gives min and max
n_distinct(gapminder$year) # gives total of unique entries
```

For `pop`
```{r}
range(gapminder$pop)
```

**Answer2**: There are data from 12 years starting in 1952 with the most recent data from 2007.
There are population sizes starting from 60011 to 1318683096 (individuals!, I guess...)


What values are typical? What’s the spread? What’s the distribution? Etc., tailored to the variable at hand.

This would be more appropriate to explore using a quantitative variable that have lots of different values, such as *pop* (population size) vs *year*.

**Answer**: We can start exploring the values of a variable using the function `summary`. In this case the mean population size for all countries in all years is 2.960e+07; the smallest population recorded is 6.001e+04, and the largest is 1.319e+09.

```{r}
summary(gapminder$pop)
```

Since there are too many values for this variable (1704 using `n_distinct(gapminder$pop)`). It would be better to see the mean population size by country across years or in a given year, for which we will have to extract subsets of data. But I guess that is the objective in the next exercise. So for now, we can just visualize the distribution of population sizes in the whole data set using an histogram

```{r}
hist(gapminder$pop)
```

Not a very nice plot, but we can see that there are many "small" values or populations and very few "big" populations.

We could also try to see this if we `cut` the variable *pop* in for example ten bins of equal size, and then count how many observations are in each bin:

```{r}
x<-cut(gapminder$pop, breaks = 10) #break up continuous variable in 10 bins
table(x)
```

Here, we can see that 1650 entries are in the lower 1/10 of all population values, which is basically between way less than 1 million and 132 millions of habitants, and very few entries are in values bigger than that...

## Explore various plot types

First, I will plot population size per year for all countries, but colouring by continent. I want to visualize both, lines and points for each country.  

```{r}
ggplot(data = gapminder, 
       aes(x=year, y=pop, 
           by=country, color=continent)) +
  geom_line() + 
  geom_point()
```

I want to see the population growth but for very large countries, so I will filter the data to visualize only countries with more than 100 million people. I'm also `piping` filtered data into ggplot in order to practice `%>%` and `filter`

```{r}
gapminder %>%
  filter(pop >= 1.00e+8) %>%
  ggplot(aes(x = year, y = pop,
             by = country, color = country)) +
  geom_point()
```

I want to also explore the distribution of population sizes by continent, so I will split continent into five panels and will use a density plot.

```{r}
ggplot(data=gapminder, 
       aes(x=pop, 
          fill=continent))+
    geom_density()+
    facet_wrap(~continent)
```

## More `filter`, `select` and `ggplot` and `%>%`

I want to see the population growth of the smallest countries in the world. Meaning countries with less than a million habitants. I will export these data into a table first using `knitr:: kable`, and then I will visualize them in a graph.  

```{r}
smallcountries <- filter(gapminder, pop <= 1.00e+6) %>%
  select(country, pop, year)
library(knitr)
kable(smallcountries[1:3])
```


Now, I want to see the population growth of the "small" countries by continent, so basically countries that have less than 1 million habitants.


```{r}
filter(gapminder, pop <= 1.00e+6) %>% 
  ggplot(aes(x=year, y=pop,
             by = country, color = country)) + 
  geom_point(alpha=0.5) + 
  facet_wrap(~continent)
```


We can see that most small countries are in Africa, and than some of them have grown to more than 1 million inhabitants in the period between 1952 and 2007.

