---
title: "Use dplyr to manipulate and explore data"
author: "Santiago D"
date: '2017-10-02'
output: github_document
---

### Load data and packages
```{r}
library("gapminder")
library("tidyverse")
```


### Tasks selected
- Maximum and minimum GDP per capita for all continents
- 

## Task #1 

**Objective**: Get the maximum and minimum of GDP per capita for all continents.

**Process**: Use `group_by` and `summarise` to obtain the min and max of the variable, save that as a new object and use that object to create a table using `knitr::kable` and plot using `ggplot2`.

**Resources**: I got some ideas for the graph from this `stackoverflow` [link](https://stackoverflow.com/questions/27205629/plot-ranges-of-variable-in-data)


```{r}
gdpsummary <- gapminder %>% 
  group_by(continent) %>% 
  summarise(min_gdpPercap = min(gdpPercap),
            max_gdpPercap = max(gdpPercap),
            mean_gdpPercap = mean(gdpPercap))
```

Table

```{r}
knitr::kable(gdpsummary, col.names = c("Continent", "Minimum GDP per capita", 
                                       "Maximum GDP per capita", "Mean GDP per capita"))
```

Figure 

```{r}
ggplot(gdpsummary, aes(x = continent)) +
  geom_point(aes(y = min_gdpPercap), size = 3, color = "skyblue4") +
  geom_point(aes(y = max_gdpPercap), size = 3, color = "skyblue4") +
  geom_linerange(aes(ymin = min_gdpPercap,ymax = max_gdpPercap),linetype = 2,color = "blue") +
  labs(title = "GDP per capita for each continent", x = "Continent", y = "GDP per capita") +
  ylim(0,120000) +
  theme_classic()
```

Optional Figure

Another option would be to use a boxplot of GDP per capita per continent, which would display a better distribution of the values, including the minimum and maximum.

```{r}
gapminder %>% 
  group_by(continent) %>% 
  ggplot(aes(x = continent, y = gdpPercap, fill = continent)) +
  geom_boxplot() +
  labs(title = "GDP per capita for each continent", x = "Continent", y = "GDP per capita") +
  ylim(0, 120000) +
  guides(fill = FALSE) +
  theme_classic()
```


## Task #2

**Objective**: How is life expectancy changing over time on different continents?

**Process**: Use `group_by` and `mutate` to create a new variable that capture the change in ife expectancy over years, but in this case by calculating the difference between life expectancy in a given year with the first year with data for each country. I saved that as a new object.

To display the change over time per continent in a table, I decided to summarize the _average_ change in life Expectancy per year and used `knitr::kable`. However for the graph, it is more informative to keep all values for each country, but including a linear regression per continent to see the trend in change of Life expectancy, of course using `ggplot2`.

```{r}
change_lifeExp <- gapminder %>% 
  group_by(country) %>% 
  arrange(year) %>% 
  mutate (change_lifeExp = lifeExp - min(lifeExp))
```

Table

```{r}
mean_change <- change_lifeExp %>% 
  group_by(continent, year) %>% 
  summarise(mean_change = mean(change_lifeExp)) 
knitr::kable(mean_change, col.names = c("Continent", "year", "Mean ∆ Life expectancy"))
```

