---
title: "Use dplyr to manipulate and explore data"
author: "Santiago D"
date: '2017-10-02'
output: github_document
---

### Load data and packages
```{r}
library("gapminder")
library("tidyverse")
```


### Tasks selected
- Maximum and minimum GDP per capita for all continents
- 

## Task #1 
**Objective**: Get the maximum and minimum of GDP per capita for all continents.
**Process**: Use `group_by` and `summarise` to obtain the min and max of the variable, save that as a new object and use that object to create a table using `knitr::kable` and plot using `ggplot2`

```{r}
gdpmaxmin <- gapminder %>% 
  group_by(continent) %>% 
  summarise(min_gdpPercap = min(gdpPercap),
            max_gdpPercap = max(gdpPercap))

knitr::kable(gdpmaxmin, col.names = c("Continent", "Minimum GDP per capita", "Maximum GDP per capita"))
```

```{r}
ggplot(gdpmaxmin) +
  geom_point(aes(x = continent, y = min_gpdPercap))
```
 working on graph....
 
```{r}
gapminder %>% 
  group_by(continent) %>% 
  summarise(min_gdpPercap = min(gdpPercap),
            max_gdpPercap = max(gdpPercap)) %>% 
  ggplot(aes(continent), colour = continent) +
  geom_point(continent, min_gdpPercap)

```

Pick at least three of the tasks below and attack each with a table and figure

