---
title: "Data Wrangling wrap up"
author: "Santiago David"
date: '2017-11-07'
output: github_document
---

#### Load data and packages
```{r}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(gapminder))
suppressPackageStartupMessages(library(singer))
suppressPackageStartupMessages(library(forcats))
suppressPackageStartupMessages(library(purrr))
suppressPackageStartupMessages(library(stringr))
data("gapminder")
```

# Writing functions

####**Objective**: 

Write one (or more) functions that do something useful to pieces of the Gapminder or Singer data. It is logical to think about computing on the mini-data frames corresponding to the data for each specific country, location, year, band, album. Make it something you can’t easily do with built-in functions. Make it something that’s not trivial to do with the simple dplyr verbs.

**Process**: I decide

Plot the data


```{r}
ggplot(gapminder, aes(year, lifeExp, colour = continent)) +
  geom_point() +
  facet_wrap(~ continent) +
  geom_smooth(method = "lm") +
  labs(title = "Life expectancy for each continent 1952-2007", x = "Years", y = "Life Expectancy (years)") +
  guides(colour = FALSE) +
  theme_classic() 
```

```{r}
summary(lm(lifeExp ~ I(year - 1952), gapminder))
```

