---
title: "Data Wrangling wrap up"
author: "Santiago David"
date: '2017-11-07'
output: github_document
---

- [Writing functions](#Writing-functions)
- [Work with a nested data frame](#Work-with-a-nested-data-frame)

#### Load data and packages
```{r}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(gapminder))
suppressPackageStartupMessages(library(purrr))
suppressPackageStartupMessages(library(stringr))
suppressPackageStartupMessages(library(broom))
suppressPackageStartupMessages(library(robust))
data("gapminder")
```

# Writing functions

**Objective**: Write one (or more) functions that do something useful to pieces of the Gapminder or Singer data. Make it something that’s not trivial to do with the simple dplyr verbs.

**Process**: The homework provide a good starting point with the gapminder data following Jenny's [tutorial](http://stat545.com/block012_function-regress-lifeexp-on-year.html) to create a linear regression of life expectancy on year. However, 
First things first, let's plot the data!

```{r}
ggplot(gapminder, aes(year, lifeExp, colour = continent)) +
  geom_point() +
  facet_wrap(~ continent) +
  labs(title = "Life expectancy for each continent 1952-2007", x = "Years", y = "Life Expectancy (years)") +
  guides(colour = FALSE) +
  theme_classic() 
```

From these graphs, we can see that there is an increasing trend of life expectancy over years, However, a priori, we dont really know if the relationship is different for each country/continent, if the relationship between Life expectancy and year is linear, cuadratic, or whether a robust regression is more apropiate given the possibility that some values are outliers or influential observations. For example, what about that steep line of points in Europe?, or that odd red dot in Africa in the 90's?. For this reason, it is worth to create functions for different models that we can fit either to each continent, or country (or other similar data).

The ideal function, will be one that takes a variable `life expectancty` and `year` as input, fit a model, and return the coefficients `intercept`, and `slope` and potentially the `R-squared` as a way to understand how well the model fit the data...

####**Create some code**

We should start by fitting the models, and check how they behave by extracting the coefficients:

```{r}
# three models
linear1 <- lm(lifeExp ~ I(year - 1952), gapminder)
quadra1 <- lm(lifeExp ~ I(year - 1952) + I((year - 1952)^2), gapminder)
robust1 <- robustbase::lmrob(lifeExp ~ I(year - 1952), gapminder)
```

We have to use `I(year - 1952)` instead of just `year` to fix the intercept to life expectancy on year 1952 (following Jenny Bryan's [post](http://stat545.com/block012_function-regress-lifeexp-on-year.html))

Now, let's see the parameters

```{r}
#coefficients
coef(linear1)
coef(quadra1)
coef(robust1)
```

And the adjusted R-squared to get a sense of how close the data are to the fitted regression line
```{r}
#adjusted r squared
summary(linear1)$adj.r.squared
summary(quadra1)$adj.r.squared
summary(robust1)$adj.r.squared
```

Now, we see that the robust and quadratic models perform a bit better than the simple linear model. But remember this is for the `whole` data. Now, we want to create a function with that code that we can apply to subsets of the data.

####**Turn code into a function**

I can fit a function for each model separately... starting with the simple linear regression

```{r}
linear_mod <- function(dat, offset = 1952) {
  model <- lm(lifeExp ~ I(year - offset), dat)
  intercept <- coef(model)[1]
  slope <- coef(model)[2]
  adjusted_Rsquared <- summary(model)$adj.r.squared
  data.frame(intercept, slope, adjusted_Rsquared) %>% 
  setNames(c("Intercept", "Slope", "R-squared"))
}
```

Note, that here, we created an `offset` to specify an argument in the function, and defined specific objects for each parameter we are interested in, by extracting the first `[1]` and second `[2]` elements of the `coef` object and the `adj.r.squared` element of the summary object. We then create a data.frame with those value... 

Lets apply this function to the whole gapminder dataset.
```{r}
linear_mod(gapminder)
```

We can see that these values correspond to the values above (model = linear1), obtained by fitting a simple regression to Life expectancy and year to gapminder. So that's good, the function is doing what we think it is doing...

**Now** We can fix a separate function for the quadractic and robust regression models, or **even better**... Try to create a function that fit all three models at once, extract the three parameters of interest `intercept`, `slope`, and `R-squared` from each model and output a single table with those values...

```{r}
fit_all_models <- function(dat, offset = 1952) {
  # three models
  linear <- lm(lifeExp ~ I(year - offset), dat)
  quadra <- lm(lifeExp ~ I(year - offset) + I((year - offset)^2), dat)
  robust <- robustbase::lmrob(lifeExp ~ I(year - offset), dat)
  # parameters
  model <- c("Linear", "Quadratic", "Robust")
  intercept <- c(coef(linear)[1], coef(quadra)[1], coef(robust)[1])
  slope <- c(coef(linear)[2], coef(quadra)[2], coef(robust)[2])
  Adjusted_Rsquared <- c(summary(linear)$adj.r.squared, summary(quadra)$adj.r.squared, 
                         summary(robust)$adj.r.squared)
  # what we want for the output
  tibble(model, intercept, slope, Adjusted_Rsquared) %>% 
  setNames(c("Model", "Intercept", "Slope", "Rsquared"))
}
```

Okay, let's see if it works for the whole gapminder... we should obtain exactly the same values we obtained in the first place, by fitting the three models to gapminder and extracting the values independently.

```{r}
fit_all_models(gapminder)
```

**Observations**: This function allows us to test three models at once to a set of data with a `Life expectancy` and `year` variables, extract the parameters of interest from each model and compared them easily in a single output table; So, now we can use this function to fit the models for each continent (or country)...

Let's try it for each continent, using the function `do`. We can check the output in a table

```{r}
gapminder %>% 
  group_by(continent) %>% 
  do(fit_all_models(.)) %>% 
  knitr::kable(format = "markdown")
```

Or visualize one parameter, such as the Adjusted R-squared per model for each continent in a graph...

```{r}
gapminder %>% 
  group_by(continent) %>% 
  do(fit_all_models(.)) %>% 
  ggplot(aes(x=continent, y= Rsquared , fill = Model)) +
           geom_bar(position = "dodge", stat = "identity") +
           theme_classic()
```

Now we can compare the `Intercept`, `Slope` and `Adjusted R-squared` per model for each continent with a single function that can be modified easily, if we want to extract something else from these models. Of course for statistical purposes, it is necessary to explore other aspects of these models, such as the estimated coefficients, F-test, etc... and not just the Adjusted R-squared values...

# Work with a nested data frame

**Objective**: Create a nested data frame and map a function over the list column holding the nested data. Try to extract parameters from the models and go back to a simple data frame with summarized results.

**Process**: 
First, I will "nest" the data in gapminder by continent and country, using `nest()`... 

```{r}
gap_nested <- gapminder %>% 
  group_by(continent, country) %>% 
  nest()
gap_nested
```

We can see that this "new" dataframe collapsed all information associated to each country in a new variable `data` of the type `list`. I don't really need to inspect this list, since we know what there is inside, but just to give an example to my future self, this is how I can access information for a specific element.

```{r}
gap_nested[[1, "data"]]
```

Now, I can start applying **my** function `fit_all_models` over the nested dataframe. Let's do it for all countries applying `purrr:map()`. Following the recommendation in the [split-apply-combine](http://stat545.com/block024_group-nest-split-map.html) lesson, we can use mutate to create a new column in the same dataframe for the output of this function...

```{r warning=FALSE}
(gap_nested <- gap_nested %>% 
  mutate(model_fits = map(data, fit_all_models)))
```

Now, we have a column with the output of the function `fit_all_models` called `model_fits`, which happens to be another tibble... Again, we can inspect the info in that element for the first country in the database...

```{r}
gap_nested[[1, "model_fits"]]
```

Finally, we can extract the information in this column for all countries by `unnesting` the dataframe to a normal tibble...

```{r}
(gap_nested %>% 
   select(continent, country, model_fits) %>% 
   unnest(model_fits) %>% 
   head())
```

**Okay**, so far so good... I practice applying a function that I created over elements of a nested dataframe. However, for this specific example, you might notice that this long process was unnecessary, given that in my original function, I already specified the elements I wanted from the models, so, just by simply applying the function over the database, grouped by continent and country, I will obtain exactly the same dataframe as above...

```{r warning=FALSE}
(gapminder %>% 
  group_by(country, continent) %>% 
  do(fit_all_models(.))) %>% 
  head() %>% 
  knitr::kable(format = "markdown")
```

A more interesting scenario would be one in which rather than obtaining just those three elements from the models, we can obtain all the information produced when you fit a linear model, such as estimates, std.errors, F-statistics, etc... All that information is usually captured in the `summary` of the object lm, and can be extracted in a nice and tidy way using the [broom](https://github.com/tidyverse/broom) package

Let's a create a new simplified function that keeps all the information from the model objects in a list object

```{r}
fit_models_complete <- function(dat, offset = 1952) {
  # three models
  linear <- lm(lifeExp ~ I(year - offset), dat)
  quadra <- lm(lifeExp ~ I(year - offset) + I((year - offset)^2), dat)
  robust <- robust::lmRob(lifeExp ~ I(year - offset), dat)
  output <- list(Linear = linear, Quadratic = quadra, Robustlm = robust)
  return(output)
} 
```

Now, just to confirm that the function is doing what we want... I'll apply it to the first country "Afghanistan", the parameters should be the same than the ones reported in the previous two tables for the `coefficients`:

```{r}
gapminder %>% 
  filter(country == "Afghanistan") %>% 
  fit_models_complete() 
```

**Note** After trying to extract information using `Broom` from a robust regression using `robustbase::lmrob`, I discovered that broom cannot deal with objects of that class... so, I have to fit a robust regression using a different method in the package `robust::lmRob`, which broom recognizes..., for that reason the robust regression parameters might be a bit different.

**Now** let's apply this function to the nested dataframe, the same way we did it before
```{r}
(gap_nested_all_models <- gap_nested %>% 
  mutate(model_fits = map(data, fit_models_complete)))
```

**Wow!!!** Now I added another level of complexity to this dataframe, the new column `model_fits`, is actually another list with 3 elements, which are the results of each model... Let's try to confirm that by inspecting the first model results for the first element (country)

```{r}
gap_nested_all_models[[1, "model_fits"]][1]
```

As we know, `lm` objects have a lot of information, and by applying the function I created, we basically calculate that for three models for each country and keep all that information in a list into the dataframe, but we need access to the information. So, let's try to use `broom`. 

```{r}
#tidy(gap_nested_all_models[[1, "model_fits"]][1])
#glance(gap_nested_all_models[[1, "model_fits"]][1])
#augment(gap_nested_all_models[[1, "model_fits"]][1])
```

It seems that the functions `tidy`, `glance`, and `augment` from **broom** need one list for one specific model at the time for each element... Given that I created a function that combine all the results from three models into one list, it is not possible to use those functions in this dataframe... OR maybe there is one?, if you have any advice... please let me know!!!, 
I tried to extract the results of one model for one element as a new object, but still it wasn't possible to apply these functions, given the structure of the "messy" dataframe I created...

The alternative would be to create one single function for each model, and incorporate the outputs to the nested dataframe, basically following the same logic for a simple linear model on the [webpage](http://stat545.com/block024_group-nest-split-map.html).
